Research log

### 2025-10-21 docs/paper.Rmd
added file readmes, added gitbook version of paper

classoption       : "man"
output            : papaja::apa6_word

@@

Having established an alignment in the total number of guesses, we next turned to the content of pretend and non-pretend guesses. In order to directly compare pretend and non-pretend guesses for the same board state, Battleship players completed two half-games in which they were instructed to continue the game from a half-completed state. In standard games, players start in the same (blank) board state but quickly diverge as they make different guess sequences. Including half games allowed us access to hundreds of cell selections for the same board state from pretenders and non-pretenders. This way, we had sufficient statistical power to compare the two guess distributions. We find a strong correlation between the spatial distributions of pretend and non-pretend guesses (board A: $r=$ `r apa_print(cor_A_pretend_nonpretend)$table$estimate`, $p$ `r apa_print(cor_A_pretend_nonpretend)$table$p.value`; board B: $r=$ `r apa_print(cor_B_pretend_nonpretend)$table$estimate`, $p$ `r apa_print(cor_B_pretend_nonpretend)$table$p.value`; Fig. \@ref(fig:decisions)C), confirming that pretenders were sensitive not only to the number of guesses they would have made had they been playing for real, but also to their content.

@@

One possible account of the reduced decision entropy in pretend games is that it reflects pretenders’ miscalibrated intuitions about randomness, conforming to a prototype of randomness that is itself too ordered. If the same prototype of randomness is used by pretenders to determine the number of unsuccessful guesses per game, the two measures should be correlated across participants. Crucially, we find the exact opposite pattern: a negative correlation between variability in the number of unsuccessful guesses and game entropy (Battleship: `r apa_print(cor.test(E2.entropy_sd$pretend_entropy,E2.entropy_sd$pretend_sd, method='spearman'))$full_result`, Hangman: `r apa_print(cor.test(E4.entropy_sd$pretend_entropy,E4.entropy_sd$pretend_sd, method='spearman'))$full_result`). This negative correlation was not observed in non-pretend games (and was even positive for Battleship: `r apa_print(cor.test(E2.entropy_sd$nonpretend_entropy,E2.entropy_sd$nonpretend_sd, method='spearman'))$full_result`; Hangman: `r apa_print(cor.test(E4.entropy_sd$nonpretend_entropy,E4.entropy_sd$nonpretend_sd, method='spearman'))$full_result`). We interpret this as evidence that the reduction in variance reflects miscalibrated beliefs not only about randomness, but also about participants’ behaviour under a counterfactual knowledge state. Those players who thought they would strictly follow a particular sequence of guesses (low entropy), ended up producing games of more variable lengths, as their success depended more on luck. Other players adjusted their decision strategy more flexibly, perhaps attempting to produce games that are not too long or short, in line with their intuitions about randomness.

>>>>

classoption       : "man"
output            : bookdown::gitbook

@@

Having established an alignment in the total number of guesses, we next turned to the content of pretend and non-pretend guesses. In order to directly compare pretend and non-pretend guesses for the same board state, Battleship players completed two half-games in which they were instructed to continue the game from a half-completed state. In standard games, players start in the same (blank) board state but quickly diverge as they make different guess sequences. Including half games allowed us access to hundreds of cell selections for the same board state from pretenders and non-pretenders. This way, we had sufficient statistical power to compare the two guess distributions. We find a strong correlation between the spatial distributions of pretend and non-pretend guesses (board A: $r=$ `r apa_print(cor_A_pretend_nonpretend)$table$estimate`, $p$ `r printp(cor_A_pretend_nonpretend$p.value, add_equals=T)`; board B: $r=$ `r apa_print(cor_B_pretend_nonpretend)$table$estimate`, $p$ `r printp(cor_B_pretend_nonpretend$p.value, add_equals=T)`; Fig. \@ref(fig:decisions)C), confirming that pretenders were sensitive not only to the number of guesses they would have made had they been playing for real, but also to their content.

@@

One possible account of the reduced decision entropy in pretend games is that it reflects pretenders’ miscalibrated intuitions about randomness, conforming to a prototype of randomness that is itself too ordered. If the same prototype of randomness is used by pretenders to determine the number of unsuccessful guesses per game, the two measures should be correlated across participants. Crucially, we find the exact opposite pattern: a negative correlation between variability in the number of unsuccessful guesses and game entropy (Battleship: `r apa_print(cor.test(E2.entropy_sd$pretend_entropy,E2.entropy_sd$pretend_sd, method='spearman'))$estimate`, $p$ `r printp(cor.test(E2.entropy_sd$pretend_entropy,E2.entropy_sd$pretend_sd, method='spearman')$p.value, add_equals=T)`, Hangman: `r apa_print(cor.test(E4.entropy_sd$pretend_entropy,E4.entropy_sd$pretend_sd, method='spearman'))$estimate`, $p$ `r printp(cor.test(E4.entropy_sd$pretend_entropy,E4.entropy_sd$pretend_sd, method='spearman')$p.value, add_equals=T)`). This negative correlation was not observed in non-pretend games (and was even positive for Battleship: `r apa_print(cor.test(E2.entropy_sd$nonpretend_entropy,E2.entropy_sd$nonpretend_sd, method='spearman'))$estimate`, $p$ `r printp(cor.test(E2.entropy_sd$nonpretend_entropy,E2.entropy_sd$nonpretend_sd, method='spearman')$p.value,add_equals=T)`; Hangman: `r apa_print(cor.test(E4.entropy_sd$nonpretend_entropy,E4.entropy_sd$nonpretend_sd, method='spearman'))$estimate`, $p$ `r printp(cor.test(E2.entropy_sd$nonpretend_entropy,E2.entropy_sd$nonpretend_sd, method='spearman')$p.value,add_equals=T)`). We interpret this as evidence that the reduction in variance reflects miscalibrated beliefs not only about randomness, but also about participants’ behaviour under a counterfactual knowledge state. Those players who thought they would strictly follow a particular sequence of guesses (low entropy), ended up producing games of more variable lengths, as their success depended more on luck. Other players adjusted their decision strategy more flexibly, perhaps attempting to produce games that are not too long or short, in line with their intuitions about randomness.
